
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Taiwan is not similar

```{r c10_001, warnings=FALSE, errors=FALSE, message=FALSE}
library(geosphere)
library(lubridate, quietly=TRUE)
library(dplyr)
library(binom)
library(ggplot2)
library(maps)
library(mapdata)
library(parallel)
library(readr)
library(plotrix)
library(tidyr)
library(maptools)
library(rvest)
Sys.setenv(TZ = "UTC") 



```

Taiwan earthquake data is available from the Seismology Center, Central Weather Bureau. It was retrieved directly into R by automatically filling out the web form to supply each months worth of data.

```{r warnings=FALSE, errors=FALSE, message=FALSE}


if(!dir.exists("../othereqdata")){
  dir.create("../othereqdata")
}
if(!file.exists("../othereqdata/eq_taiwan_raw.RData")){
  
  equrl <- 'http://scweb.cwb.gov.tw/Page.aspx?ItemId=26&loc=en&gis=n'
  eq_session <- html_session(equrl)
  eq_form <- html_form(eq_session)[[1]]
  
assemble_eq <- function(x, session_is, form_is){
  yr <- as.character(x[1])
  mnth <- formatC(x[2], width = 2, format = "d", flag = "0")
  filled_form <- set_values(form, `ctl03$ddlYear` = yr, `ctl03$ddlMonth` = mnth)
  out <- submit_form(session = eq_session, filled_form)
  eqtw <- out %>% html_nodes("#ctl03_gvEarthquake") %>%
    html_table() %>% .[[1]]
  return(eqtw)
}

frmfll <- data.frame(yrs = rep(2011:2016, each=12),mnths = rep(1:12, times=12))
eq_list <- apply(frmfll, 1, assemble_eq, session_is=eq_session, form_is=eq_form)
eq_taiwan <- bind_rows(eq_list)
names(eq_taiwan) <- c("class", "origin_time", "longitude", "latitude", "magnitude", "depth", "location")
eq_taiwan$time_UTC <- ymd_hm(eq_taiwan$origin_time) - hours(8)
eq_taiwan$depth <- as.numeric(gsub(" km","",eq_taiwan$depth))
eq_national <- eq_taiwan %>% filter(magnitude >= 0 &
                                  time_UTC >= as.POSIXct("2011-09-01T00:00:00", format="%Y-%m-%dT%H:%M:%S", tz="UTC") &
                                  time_UTC < as.POSIXct("2016-09-01T00:00:00", format="%Y-%m-%dT%H:%M:%S", tz="UTC")) %>%
  distinct()  %>% arrange(time_UTC)
#one longitude was missing the decimal sign

rm(eq_taiwan)
save(eq_national, file="../othereqdata/eq_taiwan_raw.RData")
}
```

```{r}

if(!file.exists("../othereqdata/eq_taiwan_processed.RData")){
load("../othereqdata/eq_taiwan_raw.RData")
southmost <- min(eq_national$latitude)
westmost <- min(eq_national$longitude)

eq_national <- eq_national %>% filter(
  magnitude > 0) %>% rowwise() %>% mutate(
  eq_gridpoint_y = round(distVincentyEllipsoid(c(longitude, southmost), c(longitude,latitude)) /50000,0),
  eq_gridpoint_x = round(distVincentyEllipsoid(c(westmost, latitude), c(longitude,latitude)) /50000,0),
  eq_roundedlat = destPoint(p=c(longitude, southmost), b=0, d=eq_gridpoint_y*50000)[2],
  eq_roundedlong = destPoint(p=c(westmost, eq_roundedlat), b=90, d=eq_gridpoint_x*50000)[1]) %>% ungroup()
# use maptools to calculate solar angles
sun_angles <- solarpos(matrix(c(eq_national$longitude, eq_national$latitude), ncol=2), eq_national$time_UTC)
colnames(sun_angles) <- c("eq_compass", "eq_vertical")
eq_national <- cbind(eq_national,sun_angles)
eq_national$eq_is_night <- eq_national$eq_vertical < 0
# calculate 360 degree as well as vertical
eq_national <- eq_national %>% mutate(eq_angle_360 = eq_vertical,
                                 eq_angle_360 = ifelse(eq_compass > 180, 180 - eq_angle_360, eq_angle_360),
                                 eq_angle_360 = ifelse(eq_vertical < 0 & eq_compass <= 180, 360 + eq_angle_360, eq_angle_360),
                                 eq_angle_by_10 = floor(eq_angle_360 /10) * 10)
save(eq_national, file="../othereqdata/eq_taiwan_processed.RData")
}
```

```{r}
if(!file.exists("../othereqdata/eq_taiwan_expected.RData")){
load("../othereqdata/eq_taiwan_processed.RData")

lat_range <- unique(eq_national$eq_roundedlat)
long_med <- median(eq_national$eq_roundedlong)

# 1 minute intervals for a full solar year
time1 <- ymd_hms("2015-01-01 00:00:00")
time2 <- ymd_hms("2015-12-31 23:59:00")
time_sq <- seq.POSIXt(from=time1, to=time2, by="min")


calc_angs <- function(x, longinput, timeinput){
  library(dplyr)
  sun_angles <- maptools::solarpos(matrix(c(longinput, x), ncol=2), timeinput)
  colnames(sun_angles) <- c("eq_compass", "eq_vertical")
# calculate 360 degree as well as vertical
site_summary <- as.data.frame(sun_angles) %>% mutate(eq_angle_360 = eq_vertical,
                                    eq_angle_360 = ifelse(eq_compass > 180, 180 - eq_angle_360, eq_angle_360),
                                    eq_angle_360 = ifelse(eq_vertical < 0 & eq_compass <= 180, 360 + eq_angle_360, eq_angle_360),
                                    eq_angle_by_10 = floor(eq_angle_360 /10) * 10) %>%
  group_by(eq_angle_by_10) %>% summarise(total= n())
  site_summary$lat <- x
  return(site_summary)
}
###
# Calculate the number of cores
no_cores <- detectCores() - 1
# Initiate cluster
cl <- makeCluster(no_cores)
clusterExport(cl, varlist=c("lat_range", "long_med", "time_sq", "calc_angs"))
list_angs <- parLapply(cl, lat_range,
                         function(x){calc_angs(x=x, longinput=long_med, timeinput=time_sq)})
stopCluster(cl)
###
library(tidyr)
anglong <- bind_rows(list_angs)
angwide <- spread(anglong, key=eq_angle_by_10,value=total)
rm(anglong, list_angs, time_sq)
save(angwide, file="../othereqdata/eq_taiwan_expected.RData")
}
```


```{r}
load("../othereqdata/eq_taiwan_processed.RData")
load("../othereqdata/eq_taiwan_expected.RData")
eq_night = sum(eq_national$eq_is_night)
eq_total = nrow(eq_national)
 bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
 sigmas <- c(0.682689492137086,
 0.954499736103642,
 0.997300203936740,
 0.999936657516334,
 0.999999426696856,
 0.999999998026825,
 0.999999999997440)
 lbls <- c(
 "1 sigma", "2 sigma",
 "3 sigma", "4 sigma",
 "5 sigma", "6 sigma",
 "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
old_par=par()
```

```{r}
bt <- binom.test(eq_night ,eq_total, conf.level= .999999999997440)
```

For the same period as the New Zealand data (from September 2011 to September 2016), in earthquakes data from the Taiwan there are `r eq_total` events of depth greater than 0 and magnitude greater than 0.

```{r}
feature <- c("Earliest (UTC)", "Latest (UTC)",
             "Northernmost", "Southernmost",
             "Westmost", "Eastmost",
             "Percent < Mag 3", "total entries",
             "nighttime quakes")
value <- c(as.character(min(eq_national$time_UTC)),
           as.character(max(eq_national$time_UTC)),
           as.character(max(eq_national$latitude)),
           as.character(min(eq_national$latitude)),
           as.character(min(eq_national$longitude)),
           as.character(max(eq_national$longitude)),
           as.character(round(100*sum(eq_national$magnitude < 3)/eq_total,2)),
           as.character(eq_total),
           as.character(eq_night))
data.frame(feature,value) %>% knitr::kable(caption="Data description")
```



Of the `r eq_total` in the data, `r eq_night` occurred at night, a proportion of `r round(eq_night / eq_total,4)`.  A seven sigma confidence interval for the proportion of earthquakes occurring at night is `r round(bt$conf.int[1],4)` to `r round(bt$conf.int[2],4)`, which 0.5 is well inside. A proportion of 0.5 is not even outside one confidence interval. 

```{r, fig.cap="Proportion of earthquakes at night: Taiwan. n=3353"}
### making the basic proportion graph
eq_night = sum(eq_national$eq_is_night)
eq_total = nrow(eq_national)
bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
sigmas <- c(0.682689492137086,
            0.954499736103642,
            0.997300203936740,
            0.999936657516334,
            0.999999426696856,
            0.999999998026825,
            0.999999999997440)
lbls <- c(
  "1 sigma", "2 sigma",
  "3 sigma", "4 sigma",
  "5 sigma", "6 sigma",
  "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
old_par=par()
conf_steps <- function(x, sigmas=sigmas, night=eq_night, total=eq_total){
  ci_lower <- binom.confint(night, total, method=c("wilson"), conf.level = sigmas[x])[1,5]
  ci_upper <- binom.confint(night, total, method=c("wilson"), conf.level = sigmas[x])[1,6]
  ci_data <- data.frame(step = x, ci_lower, ci_upper)
}

ci_spacing <- lapply(7:1, conf_steps, sigmas=sigmas, night=eq_night, total=eq_total)
ci_steps <- bind_rows(ci_spacing)

layout(matrix(c(1,1,1,2), ncol=4))

par(mar=c(5,6,4,2))
plot(c(min(0.5,floor(100*ci_steps[1,2])/100), max(0.5,ceiling(100*ci_steps[1,3])/100)),
       y=c(-3,8), type="n", bty="n", yaxt="n", ylab="",
       xlab="Proportion of earthquakes at night")
a <- apply(ci_steps, 1, function(x){polygon(c(x[2], x[3], x[3], x[2]), c(0, 0, 1, 1), col=bands[x[1]], border=NA)})
lines(c(.5,.5), c(0,1), col="#FFFFFF")
lines(c(.5,.5), c(0,1), lty=2, col="#777777")
lines(c(eq_night/eq_total,eq_night/eq_total), c(0,1), lwd=2)

par(mar=c(0,0,0,0))
plot(x=c(0,10), y=c(0,10), type="n", bty="n", axes=FALSE)
legend(0,5.5, legend=lbls, lty=typs, lwd=weights, col=bands, bty="n", xjust=0, title="Confidence Intervals:", y.intersp=1.1, cex=0.9)

lbls2=c("50% Night", "Actual Proportion")
typs2=c(2,1)
weights2=c(1,2)
cls2=c("#777777","#000000")
legend(0,7, legend=lbls2, lty=typs2, lwd=weights2, col=cls2, bty="n", xjust=0, title="Legend", y.intersp=1.2)



par(mar=old_par$mar)
par(mfrow=c(1,1))

```

If this collection of `r eq_total` earthquakes was all I had to test with, I would have no evidence that more earthquakes happen at night. Fortunately, I have more than 1.1 million earthquakes from other sources that indicate the general behaviour is otherwise, or I would not have much of a book.

```{r, fig.cap="Proportion of night earthquakes by magnitude,  Taiwan. n=3353"}
old_par=par()
grf <- eq_national %>% mutate(floored_mag = floor(magnitude*2)/2) %>%
  group_by(floored_mag) %>% summarise(successes = sum(eq_is_night), trials=n())

poly_conf_int <- function(success, trials, aa, stepsize, sigma, colr){
  ci <- binom.confint(success, trials, method=c("wilson"), conf.level = sigma)
  lower <- ci[1,5] 
  upper <- ci[1,6] 
  a <- polygon(x=c(aa,aa+stepsize,aa+stepsize,aa), y=c(upper,upper,lower,lower),
               col=colr, border=NA)
}


plot7sig <- function(success, trials, aa, stepsize){
  library(binom)
  #bands <- c('#ffffb2','#fed976','#feb24c','#fd8d3c','#fc4e2a','#e31a1c','#b10026')
  bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
  sigmas <- c(0.682689492137086,
              0.954499736103642,
              0.997300203936740,
              0.999936657516334,
              0.999999426696856,
              0.999999998026825,
              0.999999999997440)
  sapply(7:1, function(x){poly_conf_int(success, trials, aa, stepsize, sigmas[x], bands[x])})
  a <- lines(c(aa, aa + stepsize), c(success/trials, success/trials), lwd=2)
}


lbls <- c(
  "1 sigma", "2 sigma",
  "3 sigma", "4 sigma",
  "5 sigma", "6 sigma",
  "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
clrs <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
#clrs <- c('#ffffb2','#fed976','#feb24c','#fd8d3c','#fc4e2a','#e31a1c','#b10026')

layout(matrix(c(1,1,1,2), ncol=4))
plot(x=c(0,max(grf$floored_mag)+0.5), y=c(0,1), type="n", bty="n", xlab="Magnitude (0.5 steps)", ylab="Proportion of earthquakes at night")
a <- apply(grf,1,function(x){plot7sig(x[2],x[3],x[1],0.5)})
lines(c(0,10), c(.5,.5), col="#FFFFFF")
lines(c(0,10), c(.5,.5), lty=2, col="#777777")

par(mar=c(0,0,0,0))
plot(x=c(0,10), y=c(0,10), type="n", bty="n", axes=FALSE)
legend(0,5, legend=lbls, lty=typs, lwd=weights, col=clrs, bty="n", xjust=0, title="Confidence
Intervals:", cex=0.9)
lbls=c("Expected Proportion", "Actual Proportion")
typs=c(2,1)
weights=c(1,2)
legend(0,7, legend=lbls, lty=typs, lwd=weights, bty="n", xjust=0, title="Legend", y.intersp=1.2)


par(mar=old_par$mar)
par(mfrow=c(1,1))

```

While there is not enough data to be conclusive, it looks like for earthquakes in the 2.5 to 4.5 range (those we can be most exact about) the proportions are near to 50%. Part of the unusualness of this data set may stem from the absence of low magnitude earthquakes in the data (which often occur at night in other data sets). Below 3 magnitude earthquakes makeup under 25% of the recorded earthquakes, compared to over 90% for some of the larger collections.

```{r}
by_angle <- eq_national %>% 
  group_by(eq_angle_by_10) %>% summarise(total= n()) %>%
  mutate(daynight=ifelse(eq_angle_by_10 < 180, "day", "night"))

merged <- merge(eq_national, angwide, by.x="eq_roundedlat", by.y="lat")
agg_expected <- merged %>% select(`0`:`350`) %>% colSums(na.rm=TRUE)
expected_prop <- agg_expected / sum(agg_expected)
expected <- data.frame(eq_angle_by_10 = as.numeric(names(expected_prop)), expected_prop = as.numeric(expected_prop))
expected$expected_number = expected_prop * eq_total
act_exp <- merge(expected, by_angle, by="eq_angle_by_10", all.x=TRUE)

act_exp$total[is.na(act_exp$total)] <- 0
act_exp$daynight <- NULL
act_exp$act_prop <- act_exp$total / sum(act_exp$total)


ci_brackets <- act_exp %>% ungroup() %>% mutate(grand_total=sum(total)) %>%
  rowwise() %>% mutate(ci_lower_7 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[7])[1,5] * grand_total,
                       ci_upper_7 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[7])[1,6] * grand_total,
                       ci_lower_6 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[6])[1,5] * grand_total,
                       ci_upper_6 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[6])[1,6] * grand_total,
                       ci_lower_5 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[5])[1,5] * grand_total,
                       ci_upper_5 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[5])[1,6] * grand_total,
                       ci_lower_4 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[4])[1,5] * grand_total,
                       ci_upper_4 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[4])[1,6] * grand_total,
                       ci_lower_3 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[3])[1,5] * grand_total,
                       ci_upper_3 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[3])[1,6] * grand_total,
                       ci_lower_2 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[2])[1,5] * grand_total,
                       ci_upper_2 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[2])[1,6] * grand_total,
                       ci_lower_1 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[1])[1,5] * grand_total,
                       ci_upper_1 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[1])[1,6] * grand_total)

norm_ci <- ci_brackets

for (i in c(4,7:20)){
norm_ci[,i] <- ci_brackets[,i] - ci_brackets[,3]
}
circlesize=100
```

```{r, fig.cap="Over- and under- supply of earthquakes by angle of the sun (10 degree steps).  Taiwan. n=3353"}
norm_ci$border = 2
# need to double entries with a displacement of 10 to make each side of the item on the graph
norm_ci2 <- norm_ci
norm_ci2$eq_angle_by_10 <- norm_ci2$eq_angle_by_10  + 10
norm_ci2$border = 1
graphdata <-   bind_rows(norm_ci,norm_ci2) %>% arrange(eq_angle_by_10,border)

#### plot graph
bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
old_par=par()
layout(matrix(c(1,1,1,2), ncol=4))

# overall limits
limits=2 * max(abs(c(graphdata$ci_lower_7, graphdata$ci_upper_7)))
# plot upper confidence 7 interval using plotrix
polar.plot(graphdata$ci_upper_7, polar.pos=graphdata$eq_angle_by_10, radial.lim=c(-1*limits,limits),
           labels = "", main=NULL,lwd=0.5, rp.type="p",
           show.grid.labels=FALSE, show.grid=FALSE, mar=c(0,0,0,0),
           grid.col=bands[7], line.col=bands[7], poly.col=bands[7])
# plot upper 6 confidence interval
plot_ci_round <- function(upper_bound,x){
  polar.plot(upper_bound, polar.pos=graphdata$eq_angle_by_10, add=TRUE, radial.lim=c(-1*limits,limits),
             line.col=bands[x], lwd=0.5, rp.type="p", poly.col=bands[x])
}
plot_ci_round(graphdata$ci_upper_6, 6)
plot_ci_round(graphdata$ci_upper_5, 5)
plot_ci_round(graphdata$ci_upper_4, 4)
plot_ci_round(graphdata$ci_upper_3, 3)
plot_ci_round(graphdata$ci_upper_2, 2)
plot_ci_round(graphdata$ci_upper_1, 1)
plot_ci_round(graphdata$ci_lower_1, 2)
plot_ci_round(graphdata$ci_lower_2, 3)
plot_ci_round(graphdata$ci_lower_3, 4)
plot_ci_round(graphdata$ci_lower_4, 5)
plot_ci_round(graphdata$ci_lower_5, 6)
plot_ci_round(graphdata$ci_lower_6, 7)
polar.plot(graphdata$ci_lower_7, polar.pos=graphdata$eq_angle_by_10, add=TRUE, radial.lim=c(-1*limits,limits),
           line.col="white", lwd=0.5, rp.type="p", poly.col="white")

# plot expected guide line
polar.plot(rep(0,nrow(graphdata)), polar.pos=graphdata$eq_angle_by_10, add=TRUE,radial.lim=c(-1*limits,limits),
           rp.type="p", lty=4)
# plot 500 less than expected guide line
polar.plot(rep(-1 * circlesize,nrow(graphdata)), polar.pos=graphdata$eq_angle_by_10, add=TRUE,radial.lim=c(-1*limits,limits),
           rp.type="p", lty=1, line.col="#00000044")
# plot 500 more than expected guide line
polar.plot(rep(circlesize,nrow(graphdata)), polar.pos=graphdata$eq_angle_by_10, add=TRUE,radial.lim=c(-1*limits,limits),
           rp.type="p", lty=3, line.col="#00000044")

lines(c(-1.5,-1.2)*limits, c(0,0))
lines(c(1.5,1.2)*limits, c(0,0))
text(-1.8*limits,0, label="sunset
180", cex=0.7)
text(1.8*limits,0, label="sunrise
0", cex=0.7)


par(mar=c(0,0,0,0))
plot(x=c(0,10), y=c(0,10), type="n", bty="n", axes=FALSE, xlab="")
lbls <- c(
  "1 sigma", "2 sigma",
  "3 sigma", "4 sigma",
  "5 sigma", "6 sigma",
  "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
clrs <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
legend(0,4.5, legend=lbls, lty=typs, lwd=weights, col=clrs, bty="n", xjust=0, title="Confidence Intervals:", cex=0.9)

lbls2=c("Expected Number", paste(circlesize,"under expected"), paste(circlesize,"over expected"))
typs2=c(4,1,3)
weights2=c(1,1,1)
clrs2=c("#000000","#00000044","#00000044")
legend(0,10, legend=lbls2, lty=typs2, lwd=weights2, bty="n", xjust=0, title="Legend", y.intersp=1.2, col=clrs2)



par(mar=old_par$mar)
par(mfrow=c(1,1))


```

Unsurprisingly, given the general result, the pattern with respect to the angle of the sun is the closest to a perfect circle of all the country data sets.

## Formal Statement

Earthquakes in the Taiwan show no evidence of an increased risk of nighttime earthquakes. The Taiwanese data set does not contain the night-sensitive low magnitude earthquakes present in other data sets.

## Links {#Links15}

1 - Seismology Center, Central Weather Bureau: http://scweb.cwb.gov.tw/Default.aspx?loc=en


## Chapter Code

```{r eval=FALSE, echo=TRUE, purl=FALSE}

## ----setup, include=FALSE------------------------------------------------
knitr::opts_chunk$set(echo = FALSE)

## ----c10_001, warnings=FALSE, errors=FALSE, message=FALSE----------------
library(geosphere)
library(lubridate, quietly=TRUE)
library(dplyr)
library(binom)
library(ggplot2)
library(maps)
library(mapdata)
library(parallel)
library(readr)
library(plotrix)
library(tidyr)
library(maptools)
library(rvest)
Sys.setenv(TZ = "UTC") 




## ----warnings=FALSE, errors=FALSE, message=FALSE-------------------------


if(!dir.exists("../othereqdata")){
  dir.create("../othereqdata")
}
if(!file.exists("../othereqdata/eq_taiwan_raw.RData")){
  
  equrl <- 'http://scweb.cwb.gov.tw/Page.aspx?ItemId=26&loc=en&gis=n'
  eq_session <- html_session(equrl)
  eq_form <- html_form(eq_session)[[1]]
  
assemble_eq <- function(x, session_is, form_is){
  yr <- as.character(x[1])
  mnth <- formatC(x[2], width = 2, format = "d", flag = "0")
  filled_form <- set_values(form, `ctl03$ddlYear` = yr, `ctl03$ddlMonth` = mnth)
  out <- submit_form(session = eq_session, filled_form)
  eqtw <- out %>% html_nodes("#ctl03_gvEarthquake") %>%
    html_table() %>% .[[1]]
  return(eqtw)
}

frmfll <- data.frame(yrs = rep(2011:2016, each=12),mnths = rep(1:12, times=12))
eq_list <- apply(frmfll, 1, assemble_eq, session_is=eq_session, form_is=eq_form)
eq_taiwan <- bind_rows(eq_list)
names(eq_taiwan) <- c("class", "origin_time", "longitude", "latitude", "magnitude", "depth", "location")
eq_taiwan$time_UTC <- ymd_hm(eq_taiwan$origin_time) - hours(8)
eq_taiwan$depth <- as.numeric(gsub(" km","",eq_taiwan$depth))
eq_national <- eq_taiwan %>% filter(magnitude >= 0 &
                                  time_UTC >= as.POSIXct("2011-09-01T00:00:00", format="%Y-%m-%dT%H:%M:%S", tz="UTC") &
                                  time_UTC < as.POSIXct("2016-09-01T00:00:00", format="%Y-%m-%dT%H:%M:%S", tz="UTC")) %>%
  distinct()  %>% arrange(time_UTC)
#one longitude was missing the decimal sign

rm(eq_taiwan)
save(eq_national, file="../othereqdata/eq_taiwan_raw.RData")
}

## ------------------------------------------------------------------------

if(!file.exists("../othereqdata/eq_taiwan_processed.RData")){
load("../othereqdata/eq_taiwan_raw.RData")
southmost <- min(eq_national$latitude)
westmost <- min(eq_national$longitude)

eq_national <- eq_national %>% filter(
  magnitude > 0) %>% rowwise() %>% mutate(
  eq_gridpoint_y = round(distVincentyEllipsoid(c(longitude, southmost), c(longitude,latitude)) /50000,0),
  eq_gridpoint_x = round(distVincentyEllipsoid(c(westmost, latitude), c(longitude,latitude)) /50000,0),
  eq_roundedlat = destPoint(p=c(longitude, southmost), b=0, d=eq_gridpoint_y*50000)[2],
  eq_roundedlong = destPoint(p=c(westmost, eq_roundedlat), b=90, d=eq_gridpoint_x*50000)[1]) %>% ungroup()
# use maptools to calculate solar angles
sun_angles <- solarpos(matrix(c(eq_national$longitude, eq_national$latitude), ncol=2), eq_national$time_UTC)
colnames(sun_angles) <- c("eq_compass", "eq_vertical")
eq_national <- cbind(eq_national,sun_angles)
eq_national$eq_is_night <- eq_national$eq_vertical < 0
# calculate 360 degree as well as vertical
eq_national <- eq_national %>% mutate(eq_angle_360 = eq_vertical,
                                 eq_angle_360 = ifelse(eq_compass > 180, 180 - eq_angle_360, eq_angle_360),
                                 eq_angle_360 = ifelse(eq_vertical < 0 & eq_compass <= 180, 360 + eq_angle_360, eq_angle_360),
                                 eq_angle_by_10 = floor(eq_angle_360 /10) * 10)
save(eq_national, file="../othereqdata/eq_taiwan_processed.RData")
}

## ------------------------------------------------------------------------
if(!file.exists("../othereqdata/eq_taiwan_expected.RData")){
load("../othereqdata/eq_taiwan_processed.RData")

lat_range <- unique(eq_national$eq_roundedlat)
long_med <- median(eq_national$eq_roundedlong)

# 1 minute intervals for a full solar year
time1 <- ymd_hms("2015-01-01 00:00:00")
time2 <- ymd_hms("2015-12-31 23:59:00")
time_sq <- seq.POSIXt(from=time1, to=time2, by="min")


calc_angs <- function(x, longinput, timeinput){
  library(dplyr)
  sun_angles <- maptools::solarpos(matrix(c(longinput, x), ncol=2), timeinput)
  colnames(sun_angles) <- c("eq_compass", "eq_vertical")
# calculate 360 degree as well as vertical
site_summary <- as.data.frame(sun_angles) %>% mutate(eq_angle_360 = eq_vertical,
                                    eq_angle_360 = ifelse(eq_compass > 180, 180 - eq_angle_360, eq_angle_360),
                                    eq_angle_360 = ifelse(eq_vertical < 0 & eq_compass <= 180, 360 + eq_angle_360, eq_angle_360),
                                    eq_angle_by_10 = floor(eq_angle_360 /10) * 10) %>%
  group_by(eq_angle_by_10) %>% summarise(total= n())
  site_summary$lat <- x
  return(site_summary)
}
###
# Calculate the number of cores
no_cores <- detectCores() - 1
# Initiate cluster
cl <- makeCluster(no_cores)
clusterExport(cl, varlist=c("lat_range", "long_med", "time_sq", "calc_angs"))
list_angs <- parLapply(cl, lat_range,
                         function(x){calc_angs(x=x, longinput=long_med, timeinput=time_sq)})
stopCluster(cl)
###
library(tidyr)
anglong <- bind_rows(list_angs)
angwide <- spread(anglong, key=eq_angle_by_10,value=total)
rm(anglong, list_angs, time_sq)
save(angwide, file="../othereqdata/eq_taiwan_expected.RData")
}

## ------------------------------------------------------------------------
load("../othereqdata/eq_taiwan_processed.RData")
load("../othereqdata/eq_taiwan_expected.RData")
eq_night = sum(eq_national$eq_is_night)
eq_total = nrow(eq_national)
 bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
 sigmas <- c(0.682689492137086,
 0.954499736103642,
 0.997300203936740,
 0.999936657516334,
 0.999999426696856,
 0.999999998026825,
 0.999999999997440)
 lbls <- c(
 "1 sigma", "2 sigma",
 "3 sigma", "4 sigma",
 "5 sigma", "6 sigma",
 "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
old_par=par()

## ------------------------------------------------------------------------
bt <- binom.test(eq_night ,eq_total, conf.level= .999999999997440)

## ------------------------------------------------------------------------
feature <- c("Earliest (UTC)", "Latest (UTC)",
             "Northernmost", "Southernmost",
             "Westmost", "Eastmost",
             "Percent < Mag 3", "total entries",
             "nighttime quakes")
value <- c(as.character(min(eq_national$time_UTC)),
           as.character(max(eq_national$time_UTC)),
           as.character(max(eq_national$latitude)),
           as.character(min(eq_national$latitude)),
           as.character(min(eq_national$longitude)),
           as.character(max(eq_national$longitude)),
           as.character(round(100*sum(eq_national$magnitude < 3)/eq_total,2)),
           as.character(eq_total),
           as.character(eq_night))
data.frame(feature,value) %>% knitr::kable(caption="Data description")

## ---- fig.cap="Proportion of earthquakes at night: Taiwan. n=3353"-------
### making the basic proportion graph
eq_night = sum(eq_national$eq_is_night)
eq_total = nrow(eq_national)
bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
sigmas <- c(0.682689492137086,
            0.954499736103642,
            0.997300203936740,
            0.999936657516334,
            0.999999426696856,
            0.999999998026825,
            0.999999999997440)
lbls <- c(
  "1 sigma", "2 sigma",
  "3 sigma", "4 sigma",
  "5 sigma", "6 sigma",
  "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
old_par=par()
conf_steps <- function(x, sigmas=sigmas, night=eq_night, total=eq_total){
  ci_lower <- binom.confint(night, total, method=c("wilson"), conf.level = sigmas[x])[1,5]
  ci_upper <- binom.confint(night, total, method=c("wilson"), conf.level = sigmas[x])[1,6]
  ci_data <- data.frame(step = x, ci_lower, ci_upper)
}

ci_spacing <- lapply(7:1, conf_steps, sigmas=sigmas, night=eq_night, total=eq_total)
ci_steps <- bind_rows(ci_spacing)

layout(matrix(c(1,1,1,2), ncol=4))

par(mar=c(5,6,4,2))
plot(c(min(0.5,floor(100*ci_steps[1,2])/100), max(0.5,ceiling(100*ci_steps[1,3])/100)),
       y=c(-3,8), type="n", bty="n", yaxt="n", ylab="",
       xlab="Proportion of earthquakes at night")
a <- apply(ci_steps, 1, function(x){polygon(c(x[2], x[3], x[3], x[2]), c(0, 0, 1, 1), col=bands[x[1]], border=NA)})
lines(c(.5,.5), c(0,1), col="#FFFFFF")
lines(c(.5,.5), c(0,1), lty=2, col="#777777")
lines(c(eq_night/eq_total,eq_night/eq_total), c(0,1), lwd=2)

par(mar=c(0,0,0,0))
plot(x=c(0,10), y=c(0,10), type="n", bty="n", axes=FALSE)
legend(0,5.5, legend=lbls, lty=typs, lwd=weights, col=bands, bty="n", xjust=0, title="Confidence Intervals:", y.intersp=1.1, cex=0.9)

lbls2=c("50% Night", "Actual Proportion")
typs2=c(2,1)
weights2=c(1,2)
cls2=c("#777777","#000000")
legend(0,7, legend=lbls2, lty=typs2, lwd=weights2, col=cls2, bty="n", xjust=0, title="Legend", y.intersp=1.2)



par(mar=old_par$mar)
par(mfrow=c(1,1))


## ---- fig.cap="Proportion of night earthquakes by magnitude,  Taiwan. n=3353"----
old_par=par()
grf <- eq_national %>% mutate(floored_mag = floor(magnitude*2)/2) %>%
  group_by(floored_mag) %>% summarise(successes = sum(eq_is_night), trials=n())

poly_conf_int <- function(success, trials, aa, stepsize, sigma, colr){
  ci <- binom.confint(success, trials, method=c("wilson"), conf.level = sigma)
  lower <- ci[1,5] 
  upper <- ci[1,6] 
  a <- polygon(x=c(aa,aa+stepsize,aa+stepsize,aa), y=c(upper,upper,lower,lower),
               col=colr, border=NA)
}


plot7sig <- function(success, trials, aa, stepsize){
  library(binom)
  #bands <- c('#ffffb2','#fed976','#feb24c','#fd8d3c','#fc4e2a','#e31a1c','#b10026')
  bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
  sigmas <- c(0.682689492137086,
              0.954499736103642,
              0.997300203936740,
              0.999936657516334,
              0.999999426696856,
              0.999999998026825,
              0.999999999997440)
  sapply(7:1, function(x){poly_conf_int(success, trials, aa, stepsize, sigmas[x], bands[x])})
  a <- lines(c(aa, aa + stepsize), c(success/trials, success/trials), lwd=2)
}


lbls <- c(
  "1 sigma", "2 sigma",
  "3 sigma", "4 sigma",
  "5 sigma", "6 sigma",
  "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
clrs <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
#clrs <- c('#ffffb2','#fed976','#feb24c','#fd8d3c','#fc4e2a','#e31a1c','#b10026')

layout(matrix(c(1,1,1,2), ncol=4))
plot(x=c(0,max(grf$floored_mag)+0.5), y=c(0,1), type="n", bty="n", xlab="Magnitude (0.5 steps)", ylab="Proportion of earthquakes at night")
a <- apply(grf,1,function(x){plot7sig(x[2],x[3],x[1],0.5)})
lines(c(0,10), c(.5,.5), col="#FFFFFF")
lines(c(0,10), c(.5,.5), lty=2, col="#777777")

par(mar=c(0,0,0,0))
plot(x=c(0,10), y=c(0,10), type="n", bty="n", axes=FALSE)
legend(0,5, legend=lbls, lty=typs, lwd=weights, col=clrs, bty="n", xjust=0, title="Confidence
Intervals:", cex=0.9)
lbls=c("Expected Proportion", "Actual Proportion")
typs=c(2,1)
weights=c(1,2)
legend(0,7, legend=lbls, lty=typs, lwd=weights, bty="n", xjust=0, title="Legend", y.intersp=1.2)


par(mar=old_par$mar)
par(mfrow=c(1,1))


## ------------------------------------------------------------------------
by_angle <- eq_national %>% 
  group_by(eq_angle_by_10) %>% summarise(total= n()) %>%
  mutate(daynight=ifelse(eq_angle_by_10 < 180, "day", "night"))

merged <- merge(eq_national, angwide, by.x="eq_roundedlat", by.y="lat")
agg_expected <- merged %>% select(`0`:`350`) %>% colSums(na.rm=TRUE)
expected_prop <- agg_expected / sum(agg_expected)
expected <- data.frame(eq_angle_by_10 = as.numeric(names(expected_prop)), expected_prop = as.numeric(expected_prop))
expected$expected_number = expected_prop * eq_total
act_exp <- merge(expected, by_angle, by="eq_angle_by_10", all.x=TRUE)

act_exp$total[is.na(act_exp$total)] <- 0
act_exp$daynight <- NULL
act_exp$act_prop <- act_exp$total / sum(act_exp$total)


ci_brackets <- act_exp %>% ungroup() %>% mutate(grand_total=sum(total)) %>%
  rowwise() %>% mutate(ci_lower_7 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[7])[1,5] * grand_total,
                       ci_upper_7 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[7])[1,6] * grand_total,
                       ci_lower_6 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[6])[1,5] * grand_total,
                       ci_upper_6 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[6])[1,6] * grand_total,
                       ci_lower_5 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[5])[1,5] * grand_total,
                       ci_upper_5 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[5])[1,6] * grand_total,
                       ci_lower_4 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[4])[1,5] * grand_total,
                       ci_upper_4 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[4])[1,6] * grand_total,
                       ci_lower_3 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[3])[1,5] * grand_total,
                       ci_upper_3 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[3])[1,6] * grand_total,
                       ci_lower_2 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[2])[1,5] * grand_total,
                       ci_upper_2 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[2])[1,6] * grand_total,
                       ci_lower_1 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[1])[1,5] * grand_total,
                       ci_upper_1 = binom.confint(total, grand_total, method=c("wilson"), conf.level = sigmas[1])[1,6] * grand_total)

norm_ci <- ci_brackets

for (i in c(4,7:20)){
norm_ci[,i] <- ci_brackets[,i] - ci_brackets[,3]
}
circlesize=100

## ---- fig.cap="Over- and under- supply of earthquakes by angle of the sun (10 degree steps).  Taiwan. n=3353"----
norm_ci$border = 2
# need to double entries with a displacement of 10 to make each side of the item on the graph
norm_ci2 <- norm_ci
norm_ci2$eq_angle_by_10 <- norm_ci2$eq_angle_by_10  + 10
norm_ci2$border = 1
graphdata <-   bind_rows(norm_ci,norm_ci2) %>% arrange(eq_angle_by_10,border)

#### plot graph
bands <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
old_par=par()
layout(matrix(c(1,1,1,2), ncol=4))

# overall limits
limits=2 * max(abs(c(graphdata$ci_lower_7, graphdata$ci_upper_7)))
# plot upper confidence 7 interval using plotrix
polar.plot(graphdata$ci_upper_7, polar.pos=graphdata$eq_angle_by_10, radial.lim=c(-1*limits,limits),
           labels = "", main=NULL,lwd=0.5, rp.type="p",
           show.grid.labels=FALSE, show.grid=FALSE, mar=c(0,0,0,0),
           grid.col=bands[7], line.col=bands[7], poly.col=bands[7])
# plot upper 6 confidence interval
plot_ci_round <- function(upper_bound,x){
  polar.plot(upper_bound, polar.pos=graphdata$eq_angle_by_10, add=TRUE, radial.lim=c(-1*limits,limits),
             line.col=bands[x], lwd=0.5, rp.type="p", poly.col=bands[x])
}
plot_ci_round(graphdata$ci_upper_6, 6)
plot_ci_round(graphdata$ci_upper_5, 5)
plot_ci_round(graphdata$ci_upper_4, 4)
plot_ci_round(graphdata$ci_upper_3, 3)
plot_ci_round(graphdata$ci_upper_2, 2)
plot_ci_round(graphdata$ci_upper_1, 1)
plot_ci_round(graphdata$ci_lower_1, 2)
plot_ci_round(graphdata$ci_lower_2, 3)
plot_ci_round(graphdata$ci_lower_3, 4)
plot_ci_round(graphdata$ci_lower_4, 5)
plot_ci_round(graphdata$ci_lower_5, 6)
plot_ci_round(graphdata$ci_lower_6, 7)
polar.plot(graphdata$ci_lower_7, polar.pos=graphdata$eq_angle_by_10, add=TRUE, radial.lim=c(-1*limits,limits),
           line.col="white", lwd=0.5, rp.type="p", poly.col="white")

# plot expected guide line
polar.plot(rep(0,nrow(graphdata)), polar.pos=graphdata$eq_angle_by_10, add=TRUE,radial.lim=c(-1*limits,limits),
           rp.type="p", lty=4)
# plot 500 less than expected guide line
polar.plot(rep(-1 * circlesize,nrow(graphdata)), polar.pos=graphdata$eq_angle_by_10, add=TRUE,radial.lim=c(-1*limits,limits),
           rp.type="p", lty=1, line.col="#00000044")
# plot 500 more than expected guide line
polar.plot(rep(circlesize,nrow(graphdata)), polar.pos=graphdata$eq_angle_by_10, add=TRUE,radial.lim=c(-1*limits,limits),
           rp.type="p", lty=3, line.col="#00000044")

lines(c(-1.5,-1.2)*limits, c(0,0))
lines(c(1.5,1.2)*limits, c(0,0))
text(-1.8*limits,0, label="sunset
180", cex=0.7)
text(1.8*limits,0, label="sunrise
0", cex=0.7)


par(mar=c(0,0,0,0))
plot(x=c(0,10), y=c(0,10), type="n", bty="n", axes=FALSE, xlab="")
lbls <- c(
  "1 sigma", "2 sigma",
  "3 sigma", "4 sigma",
  "5 sigma", "6 sigma",
  "7 sigma")
typs <- c(1,1,1,1,1,1,1)
weights <- c(3,3,3,3,3,3,3)
clrs <- rev(c('#ffffcc','#d9f0a3','#addd8e','#78c679','#41ab5d','#238443','#005a32'))
legend(0,4.5, legend=lbls, lty=typs, lwd=weights, col=clrs, bty="n", xjust=0, title="Confidence Intervals:", cex=0.9)

lbls2=c("Expected Number", paste(circlesize,"under expected"), paste(circlesize,"over expected"))
typs2=c(4,1,3)
weights2=c(1,1,1)
clrs2=c("#000000","#00000044","#00000044")
legend(0,10, legend=lbls2, lty=typs2, lwd=weights2, bty="n", xjust=0, title="Legend", y.intersp=1.2, col=clrs2)



par(mar=old_par$mar)
par(mfrow=c(1,1))




```
